Estimating demographic bias on tests of children's early vocabulary

### 2022-01-18 paper/cogsci.Rmd
draft discussion

We identified 76 CDI items that function quite differently across demographic groups, and chose to prune the 15 items with large difficulty differences for both SES and race. 
After pruning these 2.2% of CDI:WS items, regressions with each demographic variable (and age, centered) showed smaller effects of SES ($\beta=0.20$ vs. $\beta=0.23$) and race ($\beta=0.48$ vs. $\beta=0.50$), while the effect of sex remained the same.

```{r demo-ability, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=5.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Density plots of language ability for children aged 23-25 months using the pruned model."}

# ToDo: maybe make these violin plots and squeeze all 6 groups side-by-side?

@@

# Note: can extract group parameters:
# coef(mod_intuitive_eth)$White$GroupPars - but means both 0, just SD varies?

@@

We investigated a popular test of children's early vocabulary for potential demographic bias, examining the distribution of words' estimated difficulties for high- vs. low-SES children, females vs. males, and for white vs. non-white children.

@@

Place acknowledgments (including funding information) in a section at
the end of the paper.

>>>>

We identified 76 CDI items that function quite differently across demographic groups, and chose to prune just the 15 items with large difficulty differences for both SES and race. 
After pruning just these 15 items from the CDI:WS and re-fitting a Rasch IRT model, regressions predicting ability with each demographic variable (and age, centered) showed reduced main effects of SES ($\beta=0.20$ vs. $\beta=0.23$) and race ($\beta=0.48$ vs. $\beta=0.50$).
Although these are small reductions in the SES and race effect sizes, we find them striking as we pruned only 2.2% of the CDI:WS items (and only 20% of the 76 extrema we identified), and did not specifically prune items that were biased against the disadvantaged groups (non-white, low-SES).

```{r demo-ability, include=F, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=5.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Density plots of language ability for children aged 23-25 months using the pruned model."}
# ToDo: maybe make these violin plots and squeeze all 6 groups side-by-side?

@@

@@

We investigated the CDI:WS, a popular parent-report measure of children's early vocabulary, for potential demographic bias, examining the distribution of words' estimated difficulties for high- vs. low-SES children, females vs. males, and for white vs. non-white children.
The IRT-based analysis revealed differential item functioning (DIF) for many items along each demographic dimension, but only in the case of sex were clear clusters of items that were more well-known to females (including feminine clothing and genitalia), and a clear item that was more well-known to males (male genitalia). 
For the rest of the items, and for SES- and race-based analysis, there was a smooth continuum of DIF, making the boundary of true DIF subjective, as this would rely on knowing the true difference in language ability between groups--which we reciprocally estimate from instruments like the CDI.
To move forward, we identified candidate DIF items by looking at the extremes of each distribution, and found that for SES and race, the majority of these extrema (66% in both cases) were easier for the majority demographic group (i.e., high-SES, white), meaning that if these items were removed, SES and race effects on language ability as measured by the CDI:WS would decrease.
In contrast, the majority of the extrema in the sex-based DIF analysis were easier for the disadvantaged group (males): if these extrema are removed, the female language advantage on the CDI:WS will likely increase.
To confirm these implications, we pruned a very small number of CDI items (15 of 680) that were DIF extrema in both SES- and race-based analyses, and found that the effect of these dimensions on language ability was reduced.

This investigation is only a first step in measuring demographic bias in the items on the CDI:WS.
In aggregate, it seems to suggest that we may be slightly overestimating the magnitude of differences in language ability along the dimensions of SES and race, and perhaps underestimating the female advantage.
However, the best way to estimate endogenous differences in ability (rather than exogenous differences, e.g. in the language environment) would be to conduct controlled, in-lab experiments measuring children's ability to learn novel words.
Future research may also look into the specific items that are easier for one group than another, and determine whether these differences can be accounted for by different environmental contexts. 
For example, it is striking that many of the extrema favoring high-SES children are animals and animal sounds (e.g., "grrr", "quack quack", "woof woof", "baa baa", "duck", "sheep", "giraffe", "zebra"): do high-SES households visit the zoo more, or do they often engage in other activities related to naming animals and noises they make?
If certain high-SES activities are driving early word learning for these children, what are the activities (and associated vocabulary) that low-SES households are instead engaging in?
A fair test of children's early vocabulary would contain a representative sample of words from all activities that children engage in, across demographic groups.

@@

[Redacted for anonymous review.]
<!--We thank members of the Language and Cognition lab for their feedback.-->
